Idealized Wind Models of Cool Gas Outflows

### 2010-08-16 Winds/papers/Theory/First/wind_simple.tex
blah
git-svn-id: https://svn.ucolick.org/xsvn/trunk/radtrans@5128 ac11df5e-a788-c23b-a8e1-c758c323924d

\newcommand{\ewabs}{$W_{\rm a}$}
\newcommand{\ewe}{$W_{\rm e}$}
\newcommand{\rtt}{$r_{\tau = 0.2}$}
\newcommand{\mrtt}{r_{\tau = 0.2}}
\newcommand{\bturb}{$b_{\rm turb}$}
\newcommand{\mbturb}{b_{\rm turb}}
\newcommand{\taud}{$\tau_{\rm dust}$}
\newcommand{\mtaud}{\tau_{\rm dust}}
\newcommand{\maconfig}{2p$^6$3s}
\newcommand{\mbconfig}{2p$^6$3p}
\newcommand{\aconfig}{a~$^6$D$^0$}
\newcommand{\zconfig}{z~$^6$D$^0$}
\newcommand{\mvr}{v_{\rm r}}
\newcommand{\naid}{\ion{Na}{1}~$\lambda\lambda 5891, 5897$}
\newcommand{\mgiid}{\ion{Mg}{2}~$\lambda\lambda 2796, 2803$}
\newcommand{\mgiia}{\ion{Mg}{2}~$\lambda 2796$}
\newcommand{\mgiib}{\ion{Mg}{2}~$\lambda 2803$}
\newcommand{\feiia}{\ion{Fe}{2}~$\lambda 2586$}
\newcommand{\feiib}{\ion{Fe}{2}~$\lambda 2600$}
\newcommand{\feiic}{\ion{Fe}{2}$^* \; \lambda 2612$}
\newcommand{\feiie}{\ion{Fe}{2}$^* \; \lambda 2626$}
\newcommand{\feiid}{\ion{Fe}{2}~$\lambda\lambda 2586, 2600$}
\newcommand{\feiis}{\ion{Fe}{2}$^*$}
\newcommand{\nmg}{$n_{\rm Mg^+}$}
\newcommand{\mnmg}{n_{\rm Mg^+}}
\newcommand{\nfe}{$n_{\rm Fe^+}$}
\newcommand{\mnfe}{n_{\rm Fe^+}}
\def\hub{h_{72}^{-1}}
\def\umfp{{\hub \, \rm Mpc}}
\def\mzq{z_q}
\def\zabs{$z_{\rm abs}$}
\def\mzabs{z_{\rm abs}}
\def\intl{\int\limits}
\def\cmma{\;\;\; ,}
\def\perd{\;\;\; .}
\def\ltk{\left [ \,}
\def\ltp{\left ( \,}
\def\ltb{\left \{ \,}
\def\rtk{\, \right  ] }
\def\rtp{\, \right  ) }
\def\rtb{\, \right \} }
\def\sci#1{{\; \times \; 10^{#1}}}
\def \rAA {\rm \AA}
\def \zem {$z_{\rm em}$}
\def \mzem {z_{\rm em}}
\def\smm{\sum\limits}
\def \cmm  {cm$^{-2}$}
\def \cmmm {cm$^{-3}$}
\def \kms  {km~s$^{-1}$}
\def \mkms  {\, {\rm km~s^{-1}}}
\def \lyaf {Ly$\alpha$ forest}
\def \Lya  {Ly$\alpha$}
\def \lya  {Ly$\alpha$}
\def \mlya  {Ly\alpha}
\def \Lyb  {Ly$\beta$}
\def \lyb  {Ly$\beta$}
\def \lyg  {Ly$\gamma$}
\def \ly5  {Ly-5}
\def \ly6  {Ly-6}
\def \ly7  {Ly-7}
\def \nhi  {$N_{\rm HI}$}
\def \mnhi  {N_{\rm HI}}
\def \lnhi {$\log N_{HI}$}
\def \mlnhi {\log N_{HI}}
\def \etal {\textit{et al.}}
\def \lyaf {Lyman--$\alpha$ forest}
\def \mnmin {\mnhi^{\rm min}}
\def \nmin {$\mnhi^{\rm min}$}
\def \O {${\mathcal O}(N,X)$}
\newcommand{\cm}[1]{\, {\rm cm^{#1}}}
\def \snrlim {SNR$_{lim}$}

\title{Simple Wind Models}

@@

Iron exhibits the most complex set of energy levels for elements
frequently studied in astrophysics.  The Fe$^+$ ion alone has over XXX
energy levels recorded \citep{iron}, and even this is an
incomplete list.  
One reason for iron's complexity is
that the majority of its configurations exhibit fine-structure splitting.
This includes the ground-state configuration (\aconfig) which is split
into 5 levels, 
labeled by the total angular momentum $J$, 
with excitation energies $T_{\rm ex} \equiv \Delta E / k$ ranging from
$T_{\rm ex} \approx 500-1500$\,K (Figure~\ref{fig:energy}).  
Transitions between these fine-structure levels are 
forbidden (magnetic-dipole) and have spontaneous decay times of several hours.

In this paper, we examine transitions between the ground-state
configuration and the energy levels of the \zconfig\
configuration.  This set of transitions (named the
UV1 multiplet) have wavelengths near 2600\AA.
There are two resonance-line transitions\footnote{We adopt the
  standard convention that a ``resonance line'' is an electric-dipole
  transition connected to the ground-state.  We also label
  non-resonant transitions with an asterisk, e.g.\ \feiic.} 
associated with this multiplet (\feiid)
corresponding to $\Delta J = 0, -1$; these are indicated by upward (black) arrows
in Figure~\ref{fig:energy}. The solid (green) downward
arrows in Figure~\ref{fig:energy} mark the transitions that are
connected to
the upper energy levels of the resonance lines.  These transitions may
occur following the absorption of a single photon by Fe$^+$ in its
ground-state.

The Figure also shows (as dashed, downward arrows) two of the
\feiis\ transitions that connect to higher energy levels of the \zconfig\
configuration.  Ignoring collisions and recombinations, these
transitions may only can only occur after the absorption
of two photons: one to raise the electron from the ground-state to an
excited state and another to raise the electron from the excited state
to one of the \zconfig\ levels with $J \le 5/2$.  The excitation of
fine-structure levels by 
the absorption of UV photons is termed indirect UV pumping
\citep[e.g][]{silva02,pcb06} and requires the ion to lie
near an intense source of UV photons (see $\S$~\ref{sec:pump}).  In the
following, we will assume that this process does not occur.

Our calculations also ignore collisional
processes\footnote{Recombination is also ignored.}, i.e.\ collisional
excitation and de-excitation of the various levels.  For the
fine-structure levels of the \aconfig\ configuration, the excitation
energies are modest ($T_{\rm ex} \sim 1000$\,K) but the critical
density $n_c$ is large.  For the \aconfig$_{9/2} \to
$\aconfig$_{7/2}$ transition, the critical density $n_e^C \approx 4
\sci{5} \cm{-3}$.  At these densities, one would predict  
detectable quantities of \ion{Fe}{1} which has not yet been observed
in galactic-scale outflows
\citep[see also][]{pcb06}.  Furthermore, observations rarely show
{\it absorption} from the
fine-structure levels of the \aconfig\ configuration and even that material
is not significantly blue-shifted \citep{rubin+10}.
In the following, we assume that electrons only occupy the
ground-state, i.e.\ the gas has zero opacity to the non-resonant
lines.
Empirically, because collisional excitation is insignificant and
then one may also neglect collisional de-excitation.  
The excitation energy for Mg$^+$ ($T_{\rm ex} \approx 50,000$\,K),  
meanwhile, is significantly higher implying 
negligible collisional processes.

@@

Nearly all of the absorption studies of galactic-scale outflows 
have focused on intensely, star-forming galaxies.  The
intrinsic emission of these galaxies is a complex combination of
stars and \ion{H}{2} regions that is then modulated by dust and gas
within the interstellar medium (ISM).  For the spectral regions studied
here, most stars show a featureless continuum but a few spectral
types do show significant \ion{Mg}{2} and \ion{Fe}{2} absorption.
[Any emission??]  \ion{H}{2} regions, meanwhile, are observed to emit
at the \mgiid\ doublet, primarily due to recombinations in the outer
layers [How about FeII?].  It is beyond the scope of this paper to
properly model the 
stellar absorption and \ion{H}{2} region emission, but the reader
should be aware that they can complicate the observed spectrum,
independent of the outflow, especially at velocities $v \approx 0 \mkms$.
In the following, we will simply assume a flat continuum 
normalized to unit value.  The size of the emitting
region $r_{\rm source}$ is a free parameter, but we tend to
restrict its value to be smaller than the minimum radial extent of any
gaseous component.

@@

>>>>

\newcommand{\ewabs}{$W_{\rm a}$}
\newcommand{\ewe}{$W_{\rm e}$}
\newcommand{\rtt}{$r_{\tau = 0.2}$}
\newcommand{\mrtt}{r_{\tau = 0.2}}
\newcommand{\bturb}{$b_{\rm turb}$}
\newcommand{\mbturb}{b_{\rm turb}}
\newcommand{\taud}{$\tau_{\rm dust}$}
\newcommand{\mtaud}{\tau_{\rm dust}}
\newcommand{\maconfig}{2p$^6$3s}
\newcommand{\mbconfig}{2p$^6$3p}
\newcommand{\aconfig}{a~$^6$D$^0$}
\newcommand{\zconfig}{z~$^6$D$^0$}
\newcommand{\mvr}{v_{\rm r}}
\newcommand{\naid}{\ion{Na}{1}~$\lambda\lambda 5891, 5897$}
\newcommand{\mgiid}{\ion{Mg}{2}~$\lambda\lambda 2796, 2803$}
\newcommand{\mgiia}{\ion{Mg}{2}~$\lambda 2796$}
\newcommand{\mgiib}{\ion{Mg}{2}~$\lambda 2803$}
\newcommand{\feiia}{\ion{Fe}{2}~$\lambda 2586$}
\newcommand{\feiib}{\ion{Fe}{2}~$\lambda 2600$}
\newcommand{\feiic}{\ion{Fe}{2}$* \; \lambda 2612$}
\newcommand{\feiie}{\ion{Fe}{2}$* \; \lambda 2626$}
\newcommand{\feiid}{\ion{Fe}{2}~$\lambda\lambda 2586, 2600$}
\newcommand{\feiis}{\ion{Fe}{2}$*$}
\newcommand{\nmg}{$n_{\rm Mg^+}$}
\newcommand{\mnmg}{n_{\rm Mg^+}}
\newcommand{\nfe}{$n_{\rm Fe^+}$}
\newcommand{\mnfe}{n_{\rm Fe^+}}
\def\hub{h_{72}^{-1}}
\def\umfp{{\hub \, \rm Mpc}}
\def\mzq{z_q}
\def\zabs{$z_{\rm abs}$}
\def\mzabs{z_{\rm abs}}
\def\intl{\int\limits}
\def\cmma{\;\;\; ,}
\def\perd{\;\;\; .}
\def\ltk{\left [ \,}
\def\ltp{\left ( \,}
\def\ltb{\left \{ \,}
\def\rtk{\, \right  ] }
\def\rtp{\, \right  ) }
\def\rtb{\, \right \} }
\def\sci#1{{\; \times \; 10^{#1}}}
\def \rAA {\rm \AA}
\def \zem {$z_{\rm em}$}
\def \mzem {z_{\rm em}}
\def\smm{\sum\limits}
\def \cmm  {cm$^{-2}$}
\def \cmmm {cm$^{-3}$}
\def \kms  {km~s$^{-1}$}
\def \mkms  {\, {\rm km~s^{-1}}}
\def \lyaf {Ly$\alpha$ forest}
\def \Lya  {Ly$\alpha$}
\def \lya  {Ly$\alpha$}
\def \mlya  {Ly\alpha}
\def \Lyb  {Ly$\beta$}
\def \lyb  {Ly$\beta$}
\def \lyg  {Ly$\gamma$}
\def \ly5  {Ly-5}
\def \ly6  {Ly-6}
\def \ly7  {Ly-7}
\def \nhi  {$N_{\rm HI}$}
\def \mnhi  {N_{\rm HI}}
\def \lnhi {$\log N_{HI}$}
\def \mlnhi {\log N_{HI}}
\def \etal {\textit{et al.}}
\def \lyaf {Lyman--$\alpha$ forest}
\def \mnmin {\mnhi^{\rm min}}
\def \nmin {$\mnhi^{\rm min}$}
\def \O {${\mathcal O}(N,X)$}
\newcommand{\cm}[1]{\, {\rm cm^{#1}}}
\def \snrlim {SNR$_{lim}$}

\title{Idealized Wind Models of Cool Gas Outflows}

@@

Iron exhibits the most complex set of energy levels for elements
frequently studied in astrophysics.  The Fe$^+$ ion alone has over
10,000 energy levels recorded \citep{iron}, and even this is an
incomplete list.  
One reason for iron's complexity is
that the majority of its configurations exhibit fine-structure splitting.
This includes the ground-state configuration (\aconfig) which is split
into 5 levels, 
labeled by the total angular momentum $J$, 
with excitation energies $T_{\rm ex} \equiv \Delta E / k$ ranging from
$T_{\rm ex} \approx 500-1500$\,K (Figure~\ref{fig:energy}).  
Transitions between these fine-structure levels are 
forbidden (magnetic-dipole) and have spontaneous decay times of several hours.

In this paper, we examine transitions between the ground-state
configuration and the energy levels of the \zconfig\
configuration.  This set of transitions (named the
UV1 multiplet) have wavelengths near 2600\AA.
There are two resonance-line transitions\footnote{We adopt the
  standard convention that a ``resonance line'' is an electric-dipole
  transition connected to the ground-state.  We also label
  non-resonant transitions with an asterisk, e.g.\ \feiic.} 
associated with this multiplet (\feiid)
corresponding to $\Delta J = 0, -1$; these are indicated by upward (black) arrows
in Figure~\ref{fig:energy}. The solid (green) downward
arrows in Figure~\ref{fig:energy} mark the non-resonant \feiis\
transitions that are connected to
the upper energy levels of the resonance lines.  These transitions may
occur following the absorption of a single photon by Fe$^+$ in its
ground-state.   This process may also be referred to as flourescence.

The Figure also shows (as dashed, downward arrows) two of the
\feiis\ transitions that connect to higher energy levels of the \zconfig\
configuration.  Ignoring collisions and recombinations, these
transitions may only occur after the absorption
of two photons: one to raise the electron from the ground-state to an
excited state and another to raise the electron from the excited state
to one of the \zconfig\ levels with $J \le 5/2$.  The excitation of
fine-structure levels by 
the absorption of UV photons is termed indirect UV pumping
\citep[e.g][]{silva02,pcb06} and requires the ion to lie
near an intense source of UV photons.  
Even a bright, star-forming galaxy emits too few photons at $\lambda
\approx 2500$\AA\ to UV-pump Fe$^+$ ions that are farther than $\sim
100$\,pc from the stars.
In the
following, we will assume that emission from this process is
negligible.

Our calculations also ignore collisional
processes\footnote{Recombination is also ignored.}, i.e.\ collisional
excitation and de-excitation of the various levels.  For the
fine-structure levels of the \aconfig\ configuration, the excitation
energies are modest ($T_{\rm ex} \sim 1000$\,K) but the critical
density $n_c$ is large.  For the \aconfig$_{9/2} \to
$\aconfig$_{7/2}$ transition, the critical density $n_e^C \approx 4
\sci{5} \cm{-3}$.  At these densities, one would predict  
detectable quantities of \ion{Fe}{1} which has not yet been observed
in galactic-scale outflows. Furthermore, observations rarely show
{\it absorption} from the
fine-structure levels of the \aconfig\ configuration and that material
is not significantly blue-shifted (Rubin et al., in prep). 
In the following, we assume that electrons only occupy the
ground-state, i.e.\ the gas has zero opacity to the non-resonant
lines.
If collisional excitation is insignificant 
then one may also neglect collisional de-excitation.  
Regarding the \ion{Mg}{2} doublet, its
excitation energy is significantly higher implying 
negligible collisional processes at essentially any density.

@@

Nearly all of the absorption studies of galactic-scale outflows 
have focused on intensely, star-forming galaxies.  The
intrinsic emission of these galaxies is a complex combination of
stars and \ion{H}{2} regions that is then modulated by dust and gas
within the interstellar medium (ISM).  For the spectral regions studied
here, most stars show a featureless continuum but a few spectral
types do show significant \ion{Mg}{2} and \ion{Fe}{2} absorption.
{\bf [Kate: Any emission??]}  \ion{H}{2} regions, meanwhile, are observed to emit
at the \mgiid\ doublet, primarily due to recombinations in the outer
layers {\bf [Kate: How about FeII?]}.  It is beyond the scope of this paper to
properly model the 
stellar absorption and \ion{H}{2} region emission, but the reader
should be aware that they can complicate the observed spectrum,
independent of the outflow, especially at velocities $v \approx 0 \mkms$.
In the following, we assume a simple flat continuum 
normalized to unit value.  The size of the emitting
region $r_{\rm source}$ is a free parameter, but we restrict its value
to be smaller than the minimum radial extent of any gaseous
component.

@@

\subsection{The Sobolev Approximation}

For a wind with a steep velocity law (i.e.\ a large gradient $dv/dr$)
and/or a a narrow intrinsic profile (i.e.\ a small Doppler parameter),
Sobolev \cite{sobolev}
